Short-time simulation of open collective quantum spin systems

### 2018-12-11 soc_squeezing/paper/collective_spin.tex
updated text with Ana Maria's edits

Collective spin systems are a versatile resource in quantum science
for a range of applications including quantum-enhanced metrology,
quantum simulation, and quantum information.  Theoretical interest in
such systems dates back to the mid-twentieth century with the
introduction of the Lipkin-Meshkov-Glick (LMG) model as a toy model
for testing many-body approximation methods in contemporary nuclear
physics\cite{lipkin1965validity, meshkov1965validity,
  glick1965validity}.  Since then, collective system systems have
additionally featured in both theoretical and experimental studies of
quantum criticality\cite{latorre2005entanglement,
  alcalde2007functional, morrison2008dissipationdriven,
  morrison2008collective, sarandy2009classical, wang2012quantum,
  kessler2012dissipative, majd2014lmg}, non-equilibrium
phenomena\cite{walls1978nonequilibrium, morrison2008dynamical,
  bhattacherjee2014nonequilibrium, klinder2015dynamical,
  maghrebi2016nonequilibrium, zhiqiang2017nonequilibrium,
  lang2018concurrence, smale2018observation}, and precision
metrology\cite{wineland1992spin, itano1993quantum,
  kitagawa1993squeezed, ma2011quantum, chen2010heralded,
  takano2009spin, appel2009mesoscopic, schleier-smith2010states,
  chen2011conditional, riedel2010atomchipbased, gross2010nonlinear,
  zhong2010simplified, agarwal1997atomic, lau2014proposal,
  huang2015quantum, wineland1992spin, kitagawa1993squeezed,
  schleier-smith2010squeezing, leroux2010implementation,
  bohnet2014reduced, cox2016deterministic, hosten2016measurement,
  hosten2016quantum, lewis-swan2018robust, norcia2018cavitymediated,
  jones2009magnetic}.  With the development of advanced AMO (atomic,
molecular, and optical) trapping, cooling, and control techniques,
collective spin systems have been experimentally realized in atomic
ensembles\cite{takano2009spin, appel2009mesoscopic, chen2010heralded,
  schleier-smith2010states, chen2011conditional}, Bose-Einstein
condensates\cite{riedel2010atomchipbased, gross2010nonlinear,
  klinder2015dynamical}, ultracold Fermi
gasses\cite{martin2013quantum, smale2018observation}, trapped
ions\cite{bohnet2016quantum}, optical
cavities\cite{leroux2010implementation, bohnet2014reduced,
  cox2016deterministic, hosten2016measurement, hosten2016quantum,
  zhiqiang2017nonequilibrium, lewis-swan2018robust,
  norcia2018cavitymediated}, and molecular
sensors\cite{jones2009magnetic}.

@@

In the absence of decoherence, permutation symmetry and total spin
conservation divide the total Hilbert space of a collective spin
system into superseletion sectors that grow only linearly with system
size $N$, thereby admitting efficient classical simulation of its
dynamics.  Decoherence will generally violate total spin conservation
and require the use of density operators, increasing the dimension of
accessible state space to $\sim N^3$\cite{hartmann2016generalized,
  xu2013simulating} and allowing for exact simulations of
$N\lesssim100$ particles.  If decoherence is sufficiently weak,
dynamics can be numerically solvable for $N\lesssim10^5$ particles via
``quantum trajectory'' Monte Carlo methods\cite{plenio1998quantumjump,
  zhang2018montecarlo} (also known as ``quantum jump'' or ``Monte
Carlo wavefunction'' methods) that can reproduce all expectation
values of interest.  When decoherence is strong, however, or when the
number of jump operators grows extensively with system size (e.g.~for
single-spin decay), these Monte Carlo methods can take a prohibitively
long time to converge, as simulations become dominated by incoherent
jumps that generate large numbers of distinct trajectories over which
to average.

@@

To benchmark our method for computing collective spin correlators, we
consider three collective spin models known to dynamically generate
spin-squeezed states: the one-axis twisting (OAT), two-axis twisting
(TAT), and twist-and-turn (TNT) models described by the collective
spin Hamiltonians\cite{kitagawa1993squeezed, micheli2003manyparticle}
\begin{align}
  \label{eq:OAT}
  H_{\t{OAT}} &= \chi S_\z^2, \\
  \label{eq:TAT}
  H_{\t{TAT}} &= \f{\chi}{3} \p{S_\z^2 - S_\y^2}, \\
  \label{eq:TNT}
  H_{\t{TNT}} &= \chi S_\z^2 + \Omega S_\x,
\end{align}
where we include a factor of $1/3$ in the TAT Hamiltonian because it
naturally appears in realistic proposals to experimentally realize
TAT\cite{liu2011spin, huang2015twoaxis}.  For simplicity, we further
fix $\Omega=S\chi$ (with $S\equiv N/2$ throughout this work) to the
critical value known to maximize the entanglement generation rate of
of TNT in the large-$N$ limit\cite{micheli2003manyparticle,
  sorelli2015fast}.  Note that the OAT model is a special case of the
zero-field Ising model, whose quantum dynamics in the presence of
decoherence via uncorrelated decay, excitation, and dephasing of
individual spins have been previously solved
analytically\cite{foss-feig2013nonequilibrium} (see Appendix
\ref{sec:OAT}, as well as the Supplementary Material of
Ref.~[\citenum{bohnet2016quantum}] for more general multi-spin
correlators).  We will benchmark against these analytical OAT results
wherever applicable.

The Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT} squeeze the
initial product state $\ket{\X}\propto\p{\ket\up+\ket\dn}^{\otimes N}$
with $S_\x\ket{\X}=S\ket{\X}$.  Our measure of spin squeezing is the
directionally-unbiased squeezing parameter determined by the maximal
gain in resolution $\Delta\phi$ of a phase angle $\phi$ over that
achieved by any spin-polarized product state
(e.g.~$\ket{\X}$)\cite{wineland1992spin, ma2011quantum},
\begin{align}
  \xi^2
  \equiv \f{\p{\Delta\phi_{\t{min}}}^2}{\p{\Delta\phi_{\t{polarized}}}^2}
  = \f{N}{\abs{\bk{\v S}}^2}
  \min_{\uv n\perp\bk{\v S}} \bk{\p{\v S\c\uv n}^2},
  \label{eq:squeezing}
\end{align}
where $\v S\equiv\p{S_\z,S_\x,S_\y}$ is a collective spin vector, and
the minimization is performed over all unit vectors $\uv n$ in the
plane orthogonal to the mean spin vector $\bk{\v S}$.  This squeezing
parameter is entirely determined by one- and two-spin correlators of
the form $\bk{S_m}$ and $\bk{S_m S_n}$.  In the case of coherent
dynamics under TAT or TNT, these correlators are obtainable via exact
simulations of quantum dynamics in the $\p{N+1}$-dimensional Dicke
manifold of states $\set{\ket{S,m}}$ with net spin $S$ and spin
projection $m$ onto the $z$ axis, i.e.~with
$\bk{S,m|\v S^2|S,m}=S\p{S+1}$ and $\bk{S,m|S_\z|S,m}=m$ for
$\abs{m}\le S$.  In the presence of permutationally-symmetric
decoherence, these correlators are obtainable with the collective-spin
quantum trajectory Monte Carlo method recently developed in
ref.~\cite{zhang2018montecarlo}.  These exact and quantum trajectory
simulations will serve as our methods for benchmarking the expansion
in Eq.~\eqref{eq:order_truncation}.

\begin{figure}
  \centering
  \includegraphics{coherent.pdf}
  \caption{Spin squeezing of $N=10^4$ spins initially in $\ket\X$
    under coherent evolution, as computed using exact methods (solid
    lines) and the expansion in Eq.~\eqref{eq:order_truncation} with
    $M=40$ (dashed lines).  Solid circles mark the times for TAT and
    TNT after which the latter method gives an unphysical result with
    $\xi^2<0$.  Both methods agree exactly in the OAT case for all
    times shown.}
  \label{fig:coherent}
\end{figure}

\begin{figure}
  \centering
  % \includegraphics{weak_decoherence.pdf}
  \caption{Spin squeezing of $N=10^4$ spins initially in $\ket\X$ with
    decoherence via uncorrelated decay, excitation, and dephasing of
    individual spins at rates $\gamma_-=\gamma_+=\gamma_\z=\chi$.
    Squeezing is computed using an exact simulation (solid line),
    quantum trajectory methods (dots), and the expansion in
    Eq.~\eqref{eq:order_truncation} with $M=40$ (dashed lines).  Solid
    circles mark the times for TAT and TNT after which the latter
    method gives an unphysical result with $\xi^2<0$.}
  \label{fig:weak_decoherence}
\end{figure}

@@

where ${ n \brack k }$ is an unsigned Stirling number of the first

@@

  = \sum_{\ell=0}^p \epsilon_{mn}^{p\ell}
  \p{-1}^\ell \p{\sp{m-p} + \bmu S_\z}^\ell

@@

  \equiv \sum_{\ell=q}^p \epsilon_{mn}^{p\ell} \p{-1}^\ell
  { \ell \choose q } \p{m-p}^{\ell-q}

@@

>>>>

Collective spin systems are a versatile resource in quantum science
for a range of applications including quantum-enhanced metrology,
quantum simulation, and quantum information.  Theoretical interest in
such systems dates back to the mid-twentieth century with the
introduction of the Dicke model\cite{dicke1954coherence} that
describes atoms cooperatively interacting with a single mode of the
radiation field and the Lipkin-Meshkov-Glick (LMG) model, a toy model
for testing many-body approximation methods in contemporary nuclear
physics\cite{lipkin1965validity, meshkov1965validity,
  glick1965validity}.  On the experimental side, the development of
advanced trapping, cooling, and control techniques in atomic,
molecular, and optical (AMO) systems have enabled the realization of
collective spin models in a broad range of platforms, including atomic
ensembles\cite{takano2009spin, appel2009mesoscopic, chen2010heralded,
  schleier-smith2010states, chen2011conditional}, Bose-Einstein
condensates\cite{riedel2010atomchipbased, gross2010nonlinear,
  klinder2015dynamical}, ultracold Fermi
gasses\cite{martin2013quantum, smale2018observation}, trapped
ions\cite{bohnet2016quantum}, optical
cavities\cite{leroux2010implementation, bohnet2014reduced,
  cox2016deterministic, hosten2016measurement, hosten2016quantum,
  zhiqiang2017nonequilibrium, lewis-swan2018robust,
  norcia2018cavitymediated}, and molecular
sensors\cite{jones2009magnetic}. These implementations have triggered
intensive investigations of rich phenomena, including studies of
quantum criticality\cite{latorre2005entanglement,
  alcalde2007functional, morrison2008dissipationdriven,
  morrison2008collective, sarandy2009classical, wang2012quantum,
  kessler2012dissipative, majd2014lmg}, non-equilibrium
phenomena\cite{walls1978nonequilibrium, morrison2008dynamical,
  bhattacherjee2014nonequilibrium, klinder2015dynamical,
  maghrebi2016nonequilibrium, zhiqiang2017nonequilibrium,
  lang2018concurrence, smale2018observation}, and precision
metrology\cite{wineland1992spin, itano1993quantum,
  kitagawa1993squeezed, ma2011quantum, takano2009spin,
  appel2009mesoscopic, chen2010heralded, schleier-smith2010states,
  chen2011conditional, riedel2010atomchipbased, gross2010nonlinear,
  zhong2010simplified, agarwal1997atomic, lau2014proposal,
  huang2015quantum, wineland1992spin, kitagawa1993squeezed,
  schleier-smith2010squeezing, leroux2010implementation,
  bohnet2014reduced, cox2016deterministic, hosten2016measurement,
  hosten2016quantum, lewis-swan2018robust, norcia2018cavitymediated,
  jones2009magnetic}.

@@

In the absence of decoherence, permutation symmetry and total spin
conservation divide the total Hilbert space of a collective spin
system into superseletion sectors that grow only linearly with system
size $N$, thereby admitting efficient classical simulation of its
dynamics.  Decoherence will generally violate total spin conservation
and require the use of density operators, increasing the dimension of
accessible state space to $\sim N^3$\cite{hartmann2016generalized,
  xu2013simulating} and allowing for exact simulations of
$N\lesssim100$ particles.  If decoherence is sufficiently weak,
dynamics can be numerically solvable for $N\lesssim10^5$ particles via
``quantum trajectory'' Monte Carlo methods\cite{plenio1998quantumjump,
  zhang2018montecarlo} (also known as ``quantum jump'' or ``Monte
Carlo wavefunction'' methods) that can reproduce all expectation
values of interest.  When decoherence is strong, however, or when the
number of jump operators grows extensively with system size (e.g.~for
single-spin decay), these Monte Carlo methods can take a prohibitively
long time to converge, as simulations become dominated by incoherent
jumps that generate large numbers of distinct trajectories that need
to be averaged.

@@

To benchmark our method for computing collective spin correlators, we
consider three collective spin models known to dynamically generate
spin-squeezed states: the one-axis twisting (OAT), two-axis twisting
(TAT), and twist-and-turn (TNT) models described by the collective
spin Hamiltonians\cite{kitagawa1993squeezed, micheli2003manyparticle}
\begin{align}
  \label{eq:OAT}
  H_{\t{OAT}} &= \chi S_\z^2, \\
  \label{eq:TAT}
  H_{\t{TAT}} &= \f{\chi}{3} \p{S_\z^2 - S_\y^2}, \\
  \label{eq:TNT}
  H_{\t{TNT}} &= \chi S_\z^2 + \Omega S_\x,
\end{align}
where we include a factor of $1/3$ in the TAT Hamiltonian because it
naturally appears in realistic proposals to experimentally realize
TAT\cite{liu2011spin, huang2015twoaxis}.  For simplicity, we further
fix $\Omega=S\chi$ (with $S\equiv N/2$ throughout this work) to the
critical value known to maximize the entanglement generation rate of
TNT in the large-$N$ limit\cite{micheli2003manyparticle,
  sorelli2015fast}.

Note that the OAT model is a special case of the zero-field Ising
model, whose quantum dynamics admits an exact analytic solution even
in the presence of decoherence\cite{foss-feig2013nonequilibrium}.  We
will benchmark our calculations using these analytical OAT results
wherever applicable (see Appendix \ref{sec:OAT}, as well as the
Supplementary Material of Ref.~[\citenum{bohnet2016quantum}]).

The Hamiltonians in Eqs.~\eqref{eq:OAT}--\eqref{eq:TNT} squeeze the
initial product state $\ket{\X}\propto\p{\ket\up+\ket\dn}^{\otimes N}$
with $S_\x\ket{\X}=S\ket{\X}$.  Our measure of spin squeezing is the
directionally-unbiased Ramsey squeezing parameter determined by the
maximal gain in resolution $\Delta\phi$ of a phase angle $\phi$ over
that achieved by any spin-polarized product state
(e.g.~$\ket{\X}$)\cite{wineland1992spin, ma2011quantum},
\begin{align}
  \xi^2
  \equiv \f{\p{\Delta\phi_{\t{min}}}^2}{\p{\Delta\phi_{\t{polarized}}}^2}
  = \f{N}{\abs{\bk{\v S}}^2}
  \min_{\uv n\perp\bk{\v S}} \bk{\p{\v S\c\uv n}^2},
  \label{eq:squeezing}
\end{align}
where $\v S\equiv\p{S_\z,S_\x,S_\y}$ is a collective spin vector, and
the minimization is performed over all unit vectors $\uv n$ in the
plane orthogonal to the mean spin vector $\bk{\v S}$.  This squeezing
parameter is entirely determined by one- and two-spin correlators of
the form $\bk{S_m}$ and $\bk{S_m S_n}$.  In the case of coherent
dynamics discussed here, these correlators are obtainable via exact
simulations of quantum dynamics in the $\p{N+1}$-dimensional Dicke
manifold of states $\set{\ket{S,m}}$ with net spin $S$ and spin
projection $m$ onto the $z$ axis, i.e.~with
$\bk{S,m|\v S^2|S,m}=S\p{S+1}$ and $\bk{S,m|S_\z|S,m}=m$ for
$m\in\set{-S,-S+1,\cdots,S}$.  In the presence of
permutationally-symmetric decoherence, these correlators are
obtainable with the collective-spin quantum trajectory Monte Carlo
method recently developed in ref.~\cite{zhang2018montecarlo}.  In this
work, these exact and quantum trajectory simulations will be used to
benchmark the expansion in Eq.~\eqref{eq:order_truncation}.

\begin{figure}
  \centering
  \includegraphics{coherent.pdf}
  \caption{Spin squeezing of $N=10^4$ spins initially in $\ket\X$
    under coherent evolution, as computed using exact methods (solid
    lines) and the expansion in Eq.~\eqref{eq:order_truncation} with
    $M=40$ (dashed lines).  Solid circles mark the times at which the
    latter method gives an unphysical result with $\xi^2<0$.  Both
    methods agree exactly in the OAT case for all times shown.}
  \label{fig:coherent}
\end{figure}

\begin{figure}
  \centering
  % \includegraphics{weak_decoherence.pdf}
  \caption{Spin squeezing of $N=10^4$ spins initially in $\ket\X$ with
    decoherence via uncorrelated decay, excitation, and dephasing of
    individual spins at rates $\gamma_-=\gamma_+=\gamma_\z=\chi$.
    Squeezing is computed using an exact simulation (solid line),
    quantum trajectory methods (dots), and the expansion in
    Eq.~\eqref{eq:order_truncation} with $M=40$ (dashed lines).  Solid
    circles mark the times at which the latter method gives an
    unphysical result with $\xi^2<0$.}
  \label{fig:weak_decoherence}
\end{figure}

@@

where ${ p \brack q }$ is an unsigned Stirling number of the first

@@

@@

  \equiv \sum_{\ell=q}^p \epsilon_{mn}^{p\ell}
  { \ell \choose q } \p{-1}^\ell \p{m-p}^{\ell-q}

@@

Here ${ p \brack s }$ is an unsigned Stirling number of the first
kind.
